---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
knitr::opts_chunk$set(echo=FALSE)
```

# Mapping Fatal Police Encounters by State

<!-- badges: start -->
<!-- badges: end -->

Which US states have the highest per capita fatal police encounters ?

```{r pkgs, echo=FALSE}
library(pacman)
p_load(dplyr, magrittr, broom, ggplot2, sf, tidyverse, maps, mapdata, 
       hrbrthemes, ggthemes, rnaturalearth, urbnmapr)
```

```{r data, echo=FALSE}
# Load Data
  # Fatal encounters data
fatal_encounters <- read.csv("Data/Fatal_Encounters.csv")
  # Geospatial data for US states
usa_states <- get_urbn_map("states", sf = TRUE)
```

```{r wrangle-data, echo=FALSE}
## Some data wrangling:

# Summarize number of fatal encounters by state
encs_by_state <- fatal_encounters %>% group_by(State) %>% summarise(n_fat_encs = n())
encs_by_state <- encs_by_state[-c(1), ] #Get rid of first row

# Add column with population for each state
encs_by_state$state_pop <- c(700000, 5000000, 3000000, 7100000, 39500000, 5700000, 
                             3600000, 700000, 1000000, 21500000, 10700000, 1500000,
                             3200000, 1800000, 12800000, 6800000, 2900000, 4500000,
                             4700000, 7000000, 6200000, 1400000, 10100000, 5700000,
                             6200000, 3000000, 1100000, 10400000, 800000, 2000000,
                             1400000, 9300000, 2100000, 3100000, 20200000, 11800000,
                             4000000, 4200000, 13000000, 1100000, 5100000, 900000,
                             6900000, 29100000, 3300000, 8600000, 600000, 7700000,
                             1800000, 5900000, 60000)

# Calc fatal encounters per 100,000 people
encs_by_state %<>% mutate(fatalencs_per_100k = (n_fat_encs / state_pop) * 100000)
```

```{r merge-data, echo=FALSE, warning=FALSE, message=FALSE}
# Set up final dataframe for plotting

# Join  number of fatal encounters by state data with geospatial data
encs_by_state_geoms <- merge(encs_by_state, usa_states, 
                             by.x = "State", by.y = "state_abbv")

# Convert new dataframe into an sf object
spatial_fatalencs <- st_as_sf(encs_by_state_geoms) #%>% st_transform(8857)
```

```{r log-fig, echo=FALSE}
# Plot
ggplot() +
  geom_sf(data = spatial_fatalencs, aes(fill = log(fatalencs_per_100k))) + #, col = "grey40"
  labs(title = "Victims of Police Violence by State",
       subtitle = "Fatal encounters per 100,000 people (logged scale)",
       fill = "Fatalities") +
  viridis::scale_fill_viridis() +
 # coord_sf(datum = NA) +
  cowplot::theme_map()
```

```{r levels-fig, echo=FALSE}
# Plot
ggplot() +
  geom_sf(data = spatial_fatalencs, aes(fill = fatalencs_per_100k)) + #, col = "grey40"
  labs(title = "Victims of Police Violence by State",
       subtitle = "Fatal encounters per 100,000 people",
       fill = "Fatalities") +
  viridis::scale_fill_viridis() +
  # coord_sf(datum = NA) +
  cowplot::theme_map()
